Multiple imputation of incomplete multilevel data using Heckman selection models

### 2023-01-24 1.Paper/Paper.Rmd
Update manuscript

---
title: Multiple imputation of incomplete multilevel data using Heckman selection models
author:
- name: Johanna Munoz*
  num: a
- name: Matthias Egger 
  num: b  
- name: Orestis Efthimiou
  num: b
- name: Vincent Audigier
  num: c  
- name: Valentijn M. T. de Jong
  num: a,d,f
- name: Thomas. P. A. Debray
  num: a,e,f
address:
- num: a
  org: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands
- num: b
  org: University of Bern, Switzerland
- num: c
  org: CNAM, Laboratoire CEDRIC-MSDMA, 2 rue Conte, 75003 Paris, France  
- num: d
  org: Data Analytics and Methods Task Force, European Medicines Agency, Amsterdam, The Netherlands
- num: e
  org: Smart Data Analysis and Statistics, Utrecht, The Netherlands 
- num: f
  org: These authors contributed equally

corres: "* Johanna Munoz. \\email{j.munozavila@umcutrecht.nl}"
presentaddress: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Universiteitsweg 100, 3584 CG Utrecht, the Netherlands
authormark: Uthor \emph{et al}.
articletype: Research article
received: 2022-11-01
revised: 2022-12-01
accepted: 2023-01-01
abstract: "Missing data is a common problem in medical research, and is commonly addressed using multiple imputation. Although traditional imputation methods allow for valid statistical inference when data are missing at random (MAR), their implementation is problematic when the presence of missingness depends on unobserved variables, i.e. the data are missing not at random (MNAR). Unfortunately, this MNAR situation is rather common, in observational studies, registries and other sources of real-world data. While several imputation methods have been proposed for addressing individual studies when data are MNAR, their application and validity in large datasets with multilevel structure remains unclear. We therefore explored the consequence of MNAR data in hierarchical data in-depth, and proposed a novel multilevel imputation method for common missing patterns in clustered datasets. This method is based on the principles of Heckman selection models and adopts a two-stage meta-analysis approach to impute binary and continuous variables that may be outcomes or predictors and that are systematically or sporadically missing. After evaluating the proposed imputation model in simulated scenarios, we illustrate it use in a cross-sectional community survey to estimate the prevalence of malaria parasitemia in children aged 2-10 years in five subregions in Uganda."

@@

Modelling data under MNAR mechanism implies to specify information about the missingness process in addition to assumptions about the observed data. Two major approach have been used to address MNAR mechanism: pattern mixture models[@little1996] and selection models [@vella1998]. One of the most popular techniques within selection models is the one proposed byHeckman (1976). [@heckman1976] Briefly, the Heckman selection model corrects for selection bias by estimating two linked equations: a outcome equation, where the missing variable is associated with predictors, and a selection equation, that accounts for the inclusion of observations in the sample. An important feature of the Heckman selection model is that it does not assume data to be MNAR, so that it can also be used when data are MCAR or MAR. It therefore offers an appealing solution to incomplete data sets when the missingness mechanism is not precisely known.

@@

Generally the same variables can be used on the matrix of predictor variables $\boldsymbol{X^O_{i}}$ and $\boldsymbol{X^S_{i}}$. However, to avoid multicollinearity problems[@puhani1997], it is recommended to include in $\boldsymbol{X^S_{i}}$ at least one variable that is not included in the outcome model. [@angrist2001] This variable is commonly known as an exclusion restriction variable (ERV), and should only be associated with the selection in the sample $r^*_{ij}$ (relevance condition) but not with the actual observation $y^*_{ij}$ (exclusion condition).[@gomes_etal20] The ERV meets by definition the relevance and exclusion conditions in order to provide independent information about the selection process and to facilitate the estimation of the Heckman model.

@@

\[\begin{pmatrix}
\epsilon^O_{ij} \\
\epsilon^S_{ij}
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\begin{pmatrix}
\sigma^2_i & \rho_i\sigma_i \\
\rho_i\sigma_i & 1
\end{pmatrix}\right)
\]

@@

Different procedures can be adopted to combine information between clusters; however, in our imputation method we opted for the two-stage approach that is often used in meta-analyses.[@simmonds2005] This is because such an approach is computationally less intensive and could potentially generate fewer convergence problems in the estimation of the Heckman hierarchical model compared to other approaches. Our method is based on the following conditional imputation model, whose parameters are $\boldsymbol{\theta}=\{\boldsymbol{\beta^O},\boldsymbol{\beta^S},\boldsymbol{\Psi^S},\boldsymbol{\Psi^O},(\sigma_i,\rho_i)_{1<=i<=N}\}$
\begin{align*}
\boldsymbol{y^*_{i}}&=\boldsymbol{X^O_{i}}(\boldsymbol{\beta^O}+b^O_{i})+\boldsymbol{\epsilon^O_{i}}\\
\boldsymbol{r^*_{i}}&=\boldsymbol{X^S_{i}}(\boldsymbol{\beta^S}+b^S_{i})+\boldsymbol{\epsilon^S_{i}}\\
b^O_{i}&\sim N(0,\boldsymbol{\Psi^O})\\
b^S_{i}&\sim N(0,\boldsymbol{\Psi^S})
\end{align*}
\[\begin{pmatrix}
\boldsymbol{\epsilon^O_{i}} \\
\boldsymbol{\epsilon^S_{i}}
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\begin{pmatrix}
\sigma^2_i & \rho_i\sigma_i \\
\rho_i\sigma_i & 1
\end{pmatrix}\right)
\]

Briefly, in a first stage  we estimate the cluster specific parameters of the Heckman model $\boldsymbol{\theta_i}=\{\boldsymbol{\hat{\beta}_i^O},\boldsymbol{\hat{\beta}_i^S},\hat{\sigma}_i,\hat{\rho}_i\}$, for all $i$ between 1 and $N$ clusters.That is, we estimate separately for each of the clusters, the parameters of the two equations in each study, the outcome model (\ref{eq:1}) and the selection model (\ref{eq:2}) via a copula model.

In a second stage, we estimate random effects multivariate meta-analysis model for the coefficient parameters. Here $\boldsymbol{\beta_i^O}$ and $\boldsymbol{\beta_i^S}$parameters are assumed to be drawn independently and identically from a latent multivariate normal distribution of parameters with a  population means $\boldsymbol{\beta^O}$ and $\boldsymbol{\beta^S}$ and population variance covariance $\boldsymbol{\psi^O}$ and $\boldsymbol{\psi^S}$ respectively. [@higgins2009].Here we assume that $\sigma_i$ and $\rho_i$ are random variables that come from a populational distribution and we estimate random effects univariate meta-analysis models for these parameters.

@@

Initially, we use the copula method to estimate the set of cluster-specific parameters, $\boldsymbol{\widehat{\theta_i}}=\boldsymbol{\{\widehat{\beta^O_i}}, \boldsymbol{\widehat{\beta^S_i}}, \widehat{\sigma_i}, \widehat{\rho_i}\}$, using all $j$ units with observable measurements $y_{ij}^{obs}$ within each cluster $i$. Here we obtain the estimates not only the parameters' point estimates $\boldsymbol{\widehat{\theta_i}}$, but also their corresponding $\boldsymbol{\widehat{S(\theta_i)}}$ within-cluster variance-covariance matrix.

@@

In this step, we pool the parameters $\boldsymbol{\widehat{\theta_i}}$ with a random effects meta-analysis model using only the clusters with observable information, i.e., those with no systematically missing outcome. In particular, we pool the $p$ coefficients of the $\boldsymbol{\beta^O}$ in the outcome equation and estimate a multivariate random effects meta-analysis model with them. Similarly we combine all $q$ coefficient parameters of the $\boldsymbol{\beta^S}$ in the selection equation. 
Thus, we can denote the study specific coefficients
$$\boldsymbol{\hat{\beta^O}_i} = \boldsymbol{\beta^O} +\boldsymbol{ b^O_i} + \boldsymbol{\epsilon'^O_{i}}$$
$$\boldsymbol{\hat{\beta^S}_i} = \boldsymbol{\beta^S} + b^S_i + \boldsymbol{\epsilon'^S_{i}}$$
using the random effects 
 $b^O_i \sim N(0,\boldsymbol{\Psi^O})$ and $b^S_i\sim N(0,\boldsymbol{\Psi^S})$ with $\epsilon'^O_{i}\sim N(0,\sigma^O_{i}(X^O_i{X^O_i}^T))$ and $\epsilon'^S_{i}\sim N(0,\sigma^S_{i}(X^S_i{X^S_i}^T)^{-1})$.

We assume that  $\sigma_i$ and $\rho_i$ are random variables coming from a population distribution, so we can express them as:
$$ log(\sigma_i) \sim N( log(\sigma),\psi^{\sigma})$$
$$ tanh^{-1}(\rho_i)\sim N( tanh^{-1}(\rho),\psi^{\rho})$$ 
For each of them, we perform a univariate random effects meta-analysis. The model for $\sigma_i$ is given by the model:
$$log(\hat\sigma_i) = \sigma + b^{\sigma}_i + \epsilon^{\sigma}_{i}$$
with $b_i^{\sigma} \sim N(0,\psi^{\sigma})$, and $\epsilon_i^{\sigma} \sim N(0,var(log(\hat{\sigma}_i)))$ and  the model for $\rho_i$ is:
$$tanh^{-1}(\hat\rho_i) = \rho + b^{\rho}_i + \epsilon^{\rho}_{i}$$
with $b_i^{\rho} \sim N(0,\psi^{\rho})$, and $\epsilon_i^{\rho} \sim N(0,var(tanh^{-1}(\hat{\rho}_i)))$.

In the case of an incomplete binary variable, the $\sigma$ parameter is not specified, so it is not necessary to perform a random effects meta-analysis model of $\sigma$ or to include it in any of the following steps in the imputation model.

@@

From the meta-analysis model, we obtain the marginal estimates $\boldsymbol{\widehat{\Theta}}=\{\widehat{\boldsymbol{\beta^S}},\widehat{\boldsymbol{\beta^O}},\hat{\sigma},\hat{\rho}\}$ and the between-cluster variance matrix $\boldsymbol{\widehat{\psi}}=\{\boldsymbol{\Psi^S},\boldsymbol{\Psi^O},\psi^{\sigma},\psi^{\rho}\}$, i.e. variance-covariance matrix of the random effects, with their corresponding variance-covariance matrices $\boldsymbol{\widehat{S_{\Theta}}}$ and $\boldsymbol{\widehat{S_{\psi}}}$, which are used to draw the $\boldsymbol{\Theta^*}$ and $\boldsymbol{\psi^*}$ parameters, from their posterior distribution as follows: [@resche-rigon2013]

\begin{align*}
\boldsymbol{\Theta^*}& \sim N(\boldsymbol{\widehat{\Theta}},\boldsymbol{\widehat{S_{\Theta}}})\\
\boldsymbol{\psi^*}& \sim N(\boldsymbol{\widehat{\psi}},\boldsymbol{\widehat{S_{\psi}}})
\end{align*}

@@

Here, we considered $X_{1i}$ and $X_{2i}$ as predictors in the outcome equation, i.e., $\boldsymbol{X_i}^O=[1,\boldsymbol{X_{1i}},\boldsymbol{X_{2i}}]$. For the selection equation we included both variables and the $X_{3i}$ exclusion restriction variable, $\boldsymbol{X_i^S}=[1,\boldsymbol{X_{1i}},\boldsymbol{X_{2i}},\boldsymbol{X_{3i}}]$. Then in case of a missing continuous variable, we calculate the latent variables  $\boldsymbol{y_{i}^*}$ and $\boldsymbol{r^*_{i}}$ as follows:
\begin{align*}
 \boldsymbol{y_{i}^*}&= \boldsymbol{\beta_i^O}\boldsymbol{X_i^O}+\boldsymbol{\epsilon^O_{i}} \\
 \boldsymbol{r_{i}^*}&= \boldsymbol{\beta_i^S}\boldsymbol{X_i^O}+\boldsymbol{\epsilon^S_{i}}
\end{align*}
Here we assumed that all coefficient parameters varied across studies, by including cluster-specific random effects as:
\begin{align*}
 \beta_{i}^O&= \beta^O + b_{i}^O\\
 \beta_{i}^S&= \beta^S + b_{i}^S
\end{align*}
We fixed coefficients $\beta^O=\{0.3,1,1\}$  and $\beta^S=\{-0.8,1.3,-0.7,1.2\}$ in order to get around 40\% of sporadically missing values on the response $y_{ij}$ in the entire data set. Additionally, we ensured that the $y_{ij}^*$ observations were systematically missing in 20% of the clusters included in the data set, by removing the outcome values in the 20% of the clusters.

We assumed that random effects were independent within equations ($b_{0}^O\indep b_{1}^O \indep b_{2}^O$ and $b_{0}^S \indep b_{1}^S \indep b_{2}^S\indep b_{3}^S$), but were linked between both selection and outcome equations through a bivariate normal distributed as: 
\[\begin{pmatrix}
b^O \\
b^S
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},0.4^2\begin{pmatrix}
1 & \rho*0.4 \\
\rho*0.4 &1 
\end{pmatrix}\right)
\]
for the coefficients of intercept,$X_1$,$X_2$ and $X_3$.  We considered that the correlation parameter of the random effects between equations is 40% of the value of the assumed correlation parameter between error terms $\rho$. In addition, we included a random effect on the exclusion restriction variable given by $b_3\sim N(0,0.2^2)$ assuming that the intracluster variation in the exclusion restriction effect is lower than the variation on other coefficient parameters effects. The $\rho$ parameter was given different values depending on the simulated missing mechanism (Table \ref{tab:scenarios}) .

As regards the error terms, they were bivariate normal distributed as:
\[\begin{pmatrix}
\epsilon^O_{i} \\
\epsilon^S_{i}
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\begin{pmatrix}
\sigma^2_i & \rho\sigma_i \\
\rho\sigma_i & 1
\end{pmatrix}\right)
\]
whose $\sigma_i^2$ is a random parameter across clusters distributed as $log(\sigma_i) \sim N(0,0.05)$.

@@

## Estimands
The estimands were the parameter coefficients of the outcome equation $\beta^O=\{\beta^O_0,\beta^O_1,\beta^O_2\}$, with special emphasis on the treatment effect parameter $\beta^0_1$. We also report the estimated standard deviation from the covariance matrix of the random effects, i.e.,  $\sqrt{\psi_{00}}$,$\sqrt{\psi_{11}}$,$\sqrt{\psi_{22}}$.

@@

After the imputation procedure, we estimated the following (generalized) mixed linear effect model using the *lmer()* function from the *lme4* R package.[@bates2022]
$y_{i}=\beta_i^OX_i^O+\epsilon^O_{i}$
In case of missing binary variable, we used the same matrix of predictors but on a binary model estimated with the *glmer()* function from the *lme4* R package. Then, we pooled the estimates of the $\beta_i^O$ and the variance of the random effect and residual errors of the multiple imputed datasets according to Rubin's rule[@rubin1987], over which we calculated the performance measures on the estimands.

@@

The Heckman 2L imputation method is available in the micemd R package [@audigier2021] (as \textbf{mice.2l.2stage.heckman()}) and also on the github repository \url{https://github.com/johamunoz/Statsmed_Heckman} where you can also find all the codes accompanying this paper and a toy example that explains how to implement the method in mice.

@@

Figure \ref{fig:rhocont} shows the results of simulations where the missing variable was continuous. In the MAR scenario, i.e. when $\rho=0$, all imputation methods provided similar unbiased estimates of the coefficient parameters, but as rho increased, i.e. the mechanism became MNAR, the estimates for the complete case analysis and the MAR-IPD imputation method became biased.

 ```{r rhocont,fig.align='center',fig.pos="!htb",fig.cap="Continuous incomplete variable by varing $\\rho$, where dashed lines depict the target performance criteria value",echo=FALSE,fig.keep="last",fig.height = 3.9, fig.width = 6}
plot_rho
 ```
As it is expected both Heckman-based imputation (1l. and 2l.) models provided less biased estimates of the coefficient parameters in MNAR scenarios, but the random effects estimates on the 1l.heckman were far away from the true values as no cluster information was considered at all.

Regarding coverage, the 2l.Heckman imputation method provided the best coverage values across all the $\rho$ scenarios with a coverage level close to the nominal 95\% interval. Even though 2l.Heckman was not properly a randomization-valid method [@buuren18c], as it was not unbiased and had a coverage above 95\% across all the $\rho$ values, it led to better results in terms of bias and coverage compared to the evaluated methods.

The 2l.Heckman method, also resulted in better estimates in terms of RMSE than the other methods evaluated, via a better trade-off between bias and variance.

In particular, our method provided an advantage over the 1l.Heckman method in systematic missingness scenarios, by adding a cluster variable in the imputation model, a feature lacking from 1l.Heckman. This can be seen in the estimates of the random effects parameters of the 1l.Heckman method, which were more biased than those of the other methods in which cluster information was included, i.e. 2l.MAR and 2l.Heckman.

>>>>

---
title: Multiple imputation of incomplete multilevel data using Heckman selection models
author:
- name: Johanna Muñoz*
  num: a
- name: Orestis Efthimiou
  num: b
- name: Vincent Audigier
  num: c  
- name: Valentijn M. T. de Jong
  num: a,d,f
- name: Thomas. P. A. Debray
  num: a,e,f
address:
- num: a
  org: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands
- num: b
  org: University of Bern, Switzerland
- num: c
  org: CNAM, Laboratoire CEDRIC-MSDMA, 2 rue Conte, 75003 Paris, France  
- num: d
  org: Data Analytics and Methods Task Force, European Medicines Agency, Amsterdam, The Netherlands
- num: e
  org: Smart Data Analysis and Statistics, Utrecht, The Netherlands 
- num: f
  org: These authors contributed equally

corres: "* Johanna Munoz. \\email{j.munozavila@umcutrecht.nl}"
presentaddress: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Universiteitsweg 100, 3584 CG Utrecht, the Netherlands
authormark: Uthor \emph{et al}.
articletype: Research article
abstract: "Missing data is a common problem in medical research, and is commonly addressed using multiple imputation. Although traditional imputation methods allow for valid statistical inference when data are missing at random (MAR), their implementation is problematic when the presence of missingness depends on unobserved variables, i.e. the data are missing not at random (MNAR). Unfortunately, this MNAR situation is rather common, in observational studies, registries and other sources of real-world data. While several imputation methods have been proposed for addressing individual studies when data are MNAR, their application and validity in large datasets with multilevel structure remains unclear. We therefore explored the consequence of MNAR data in hierarchical data in-depth, and proposed a novel multilevel imputation method for common missing patterns in clustered datasets. This method is based on the principles of Heckman selection models and adopts a two-stage meta-analysis approach to impute binary and continuous variables that may be outcomes or predictors and that are systematically or sporadically missing. After evaluating the proposed imputation model in simulated scenarios, we illustrate it use in a cross-sectional community survey to estimate the prevalence of malaria parasitemia in children aged 2-10 years in five subregions in Uganda."

@@

Modelling data under MNAR mechanism implies to specify information about the missingness process in addition to assumptions about the observed data. Two major approach have been used to address MNAR mechanism: pattern mixture models[@little1996] and selection models [@vella1998]. One of the most popular techniques within selection models is the one proposed by Heckman (1976). [@heckman1976] Briefly, the Heckman selection model corrects for selection bias by estimating two linked equations: a outcome equation, where the missing variable is associated with predictors, and a selection equation, that accounts for the inclusion of observations in the sample. An important feature of the Heckman selection model is that it does not assume data to be MNAR, so that it can also be used when data are MCAR or MAR. It therefore offers an appealing solution to incomplete data sets when the missingness mechanism is not precisely known.

@@

Generally the same variables can be used on the matrix of predictor variables $\boldsymbol{X^O_{i}}$ and $\boldsymbol{X^S_{i}}$. However, to avoid multicollinearity problems[@puhani1997], it is recommended to include in $\boldsymbol{X^S_{i}}$ at least one variable that is not included in the outcome model. [@angrist2001] This variable is commonly known as an exclusion restriction variable (ERV), and should only be associated with the selection in the sample $\boldsymbol{r^*_{i}}$ (relevance condition) but not with the actual observation $\boldsymbol{y^*_{i}}$ (exclusion condition).[@gomes_etal20] The ERV meets by definition the relevance and exclusion conditions in order to provide independent information about the selection process and to facilitate the estimation of the Heckman model.

@@

\[\begin{pmatrix}
\boldsymbol{\epsilon^O_{i}} \\
\boldsymbol{\epsilon^S_{i}}
\end{pmatrix}\sim N_{n_i}\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\begin{pmatrix}
\sigma^2_i & \rho_i\sigma_i \\
\rho_i\sigma_i & 1
\end{pmatrix}\right)
\]

@@

Different procedures can be adopted to combine information between clusters; however, in our imputation method we opted for the two-stage approach that is often used in meta-analyses.[@simmonds2005] This is because such an approach is computationally less intensive and could potentially generate fewer convergence problems in the estimation of the Heckman hierarchical model compared to other approaches. Our method is based on the following conditional imputation model, whose parameters are $\boldsymbol{\theta}=\{\boldsymbol{\beta^O},\boldsymbol{\beta^S},\boldsymbol{\Psi^S},\boldsymbol{\Psi^O},(\sigma_i,\rho_i)_{1<=i<=N}\}$
\begin{align*}
\boldsymbol{y^*_{i}}&=\boldsymbol{X^O_{i}}(\boldsymbol{\beta^O}+b^O_{i})+\boldsymbol{\epsilon^O_{i}}\\
\boldsymbol{r^*_{i}}&=\boldsymbol{X^S_{i}}(\boldsymbol{\beta^S}+b^S_{i})+\boldsymbol{\epsilon^S_{i}}\\
b^O_{i}&\sim N_p(0,\boldsymbol{\Psi^O})\\
b^S_{i}&\sim N_q(0,\boldsymbol{\Psi^S})
\end{align*}
\[\begin{pmatrix}
\boldsymbol{\epsilon^O_{i}} \\
\boldsymbol{\epsilon^S_{i}}
\end{pmatrix}\sim N_{n_i}\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\begin{pmatrix}
\sigma^2_i & \rho_i\sigma_i \\
\rho_i\sigma_i & 1
\end{pmatrix}\right)
\]

Briefly, in a first stage  we estimate the cluster specific parameters of the Heckman model $\widehat{\boldsymbol{\theta_i}}=\boldsymbol{\{\widehat{\beta^O_i}}, \boldsymbol{\widehat{\beta^S_i}}, \widehat{\sigma_i}, \widehat{\rho_i}\}$, for all $i$ between 1 and $N$ clusters. That is, we estimate separately for each of the clusters, the parameters of the two equations in each study, the outcome model (\ref{eq:1}) and the selection model (\ref{eq:2}) via a copula model.

In a second stage, we estimate random effects multivariate meta-analysis model for the coefficient parameters. Here $\boldsymbol{\beta_i^O}$ and $\boldsymbol{\beta_i^S}$parameters are assumed to be drawn independently and identically from a latent multivariate normal distribution of parameters with a  population means $\boldsymbol{\beta^O}$ and $\boldsymbol{\beta^S}$ and population variance covariance $\boldsymbol{\psi^O}$ and $\boldsymbol{\psi^S}$ respectively. [@higgins2009] Here we assume that $\sigma_i$ and $\rho_i$ are random variables that come from a populational distribution and we estimate random effects univariate meta-analysis models for these parameters.

@@

Initially, we use the copula method to estimate the set of cluster-specific parameters, $\widehat{\boldsymbol{\theta_i}}=\boldsymbol{\{\widehat{\beta^O_i}}, \boldsymbol{\widehat{\beta^S_i}}, \widehat{\sigma_i}, \widehat{\rho_i}\}$, using all $j$ units with observable measurements $y_{ij}^{obs}$ within each cluster $i$. Here we obtain the estimates not only the parameters' point estimates $\widehat{\boldsymbol{\theta_i}}$, but also their corresponding $\boldsymbol{\widehat{S(\theta_i)}}$ within-cluster variance-covariance matrix.

@@

In this step, we pool the parameters $\widehat{\boldsymbol{\theta_i}}$ with a random effects meta-analysis model using only the clusters with observable information, i.e., those with no systematically missing outcome. In particular, we pool the $p$ coefficients of the $\boldsymbol{\beta^O}$ in the outcome equation and estimate a multivariate random effects meta-analysis model with them. Similarly we combine all $q$ coefficient parameters of the $\boldsymbol{\beta^S}$ in the selection equation. 
Thus, we can denote the study specific coefficients
$$\boldsymbol{\widehat{\beta^O_i}} = \boldsymbol{\beta^O} +\boldsymbol{ b^O_i} + \boldsymbol{\epsilon'^O_{i}}$$
$$\boldsymbol{\widehat{\beta^S_i}} = \boldsymbol{\beta^S} + b^S_i + \boldsymbol{\epsilon'^S_{i}}$$
using the random effects 
 $b^O_i \sim N(0,\boldsymbol{\Psi^O})$ and $b^S_i\sim N(0,\boldsymbol{\Psi^S})$ with sampling errors $\epsilon'^O_{i}$ and $\epsilon'^S_{i}$.

We assume that  $\sigma_i$ and $\rho_i$ are random variables coming from a population distribution, so we can express them as:
$$ log(\sigma_i) \sim N( log(\sigma),\psi^{\sigma})$$
$$ tanh^{-1}(\rho_i)\sim N( tanh^{-1}(\rho),\psi^{\rho})$$ 
For each of them, we perform a univariate random effects meta-analysis. The model for $\sigma_i$ is given by the model:

$$log(\widehat{\sigma}_i)= \sigma + b^{\sigma}_i + \epsilon^{\sigma}_{i}$$

with $b_i^{\sigma} \sim N(0,\psi^{\sigma})$, and $\epsilon_i^{\sigma} \sim N(0,var(log(\widehat{\sigma_i})))$.
In the case of an incomplete binary variable, the $\sigma_i$ parameter is not specified, so it is not necessary to perform a random effects meta-analysis model of $\sigma_i$ or to include it in any of the following steps in the imputation model.

The model for $\rho_i$ is given by:
$$tanh^{-1}(\widehat{\rho_i}) = \rho + b^{\rho}_i + \epsilon^{\rho}_{i}$$
with $b_i^{\rho} \sim N(0,\psi^{\rho})$, and $\epsilon_i^{\rho} \sim N(0,var(tanh^{-1}(\widehat{\rho_i})))$.

@@

From the meta-analysis model, we obtain the marginal estimates
$\widehat{\boldsymbol{\Theta}}=\boldsymbol{\{\widehat{\beta^O}}, \boldsymbol{\widehat{\beta^S}}, \widehat{\sigma}, \widehat{\rho}\}$
and the between-cluster variance matrix
$\widehat{\boldsymbol{\psi}}=\boldsymbol{\{\widehat{\Psi^O}}, \boldsymbol{\widehat{\Psi^S}}, \widehat{\psi^{\sigma}}, \widehat{\psi^{\rho}}\}$, i.e. variance-covariance matrix of the random effects, with their corresponding variance-covariance matrices $\boldsymbol{\widehat{S_{\Theta}}}$ and $\boldsymbol{\widehat{S_{\psi}}}$, which are used to draw the $\boldsymbol{\Theta^*}$ and $\boldsymbol{\psi^*}$ parameters, from their posterior distribution as follows: [@resche-rigon2013]

\begin{align*}
\boldsymbol{\Theta^*}& \sim N(\widehat{\boldsymbol{\Theta}},\widehat{\boldsymbol{S_{\Theta}}})\\
\boldsymbol{\psi^*}& \sim N(\widehat{\boldsymbol{\psi}},\widehat{\boldsymbol{S_{\psi}}})
\end{align*}

@@

Here, we considered $X_{1i}$ and $X_{2i}$ as predictors in the outcome equation, i.e., $\boldsymbol{X_i}^O=[1,\boldsymbol{X_{1i}},\boldsymbol{X_{2i}}]$. For the selection equation we included both variables and the $X_{3i}$ exclusion restriction variable, $\boldsymbol{X_i^S}=[1,\boldsymbol{X_{1i}},\boldsymbol{X_{2i}},\boldsymbol{X_{3i}}]$. Then in case of a missing continuous variable, we calculate the latent variables  $\boldsymbol{y_{i}^*}$ and $\boldsymbol{r^*_{i}}$ as follows:
\begin{align*}
 \boldsymbol{y_{i}^*}&= \boldsymbol{X_i^O}\boldsymbol{\beta_i^O}+\boldsymbol{\epsilon^O_{i}} \\
 \boldsymbol{r_{i}^*}&= \boldsymbol{X_i^O}\boldsymbol{\beta_i^S}+\boldsymbol{\epsilon^S_{i}}
\end{align*}
where $\boldsymbol{\beta_i^O}=(\beta_{i0}^O,\beta_{i1}^O,\beta_{i2}^O)^T$ and $\boldsymbol{\beta_i^S}=(\beta_{i0}^S,\beta_{i1}^S,\beta_{i2}^S,\beta_{i3}^S)^T$. Here we assumed that the coefficient parameter k varied across studies, by including cluster-specific random effects as:
\begin{align*}
 \beta_{ik}^O&= \beta_k^O + b_{ik}^O\\
 \beta_{ik}^S&= \beta_k^S + b_{ik}^S
\end{align*}
Denoting $\boldsymbol{\beta^O}$=$(\beta_{0}^O,\beta_{1}^O,\beta_{2}^O)^T$ and $\boldsymbol{\beta^S}$=$(\beta_{0}^S,\beta_{1}^S,\beta_{2}^S,\beta_{3}^S)^T$, we fixed coefficients $\boldsymbol{\beta^O}$=$(0.3,1,1)$  and $\boldsymbol{\beta^S}$=$(-0.8,1.3,-0.7,1.2)$ in order to get around 40\% of sporadically missing values on the response $y_{ij}$ in the entire data set. Additionally, we ensured that the $y_{ij}^*$ observations were systematically missing in 20% of the clusters included in the data set, by removing the outcome values in the 20% of the clusters.

We assumed that random effects were independent within equations ($b_{i0}^O\indep b_{i1}^O \indep b_{i2}^O$ and $b_{i0}^S \indep b_{i1}^S \indep b_{i2}^S\indep b_{i3}^S$), but were linked between both selection and outcome equations through a bivariate normal distributed as: 
\[\begin{pmatrix}
b_{ik}^O \\
b_{ik}^S
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\begin{pmatrix}
{\psi^O_{kk}} & \rho_b\sqrt{\psi^O_{kk}}\sqrt{\psi^S_{kk}} \\
\rho_b \sqrt{\psi^O_{kk}}\sqrt{\psi^S_{kk}} &{\psi^S_{kk}}
\end{pmatrix}\right)
\]
with ${\psi^S_{00}}={\psi^O_{00}}={\psi^S_{11}}={\psi^O_{11}}={\psi^S_{22}}={\psi^O_{22}}=0.4$ and $\rho_b=\rho*0.4$. We considered that the correlation parameter of the random effects between equations is 40% of the value of the assumed correlation parameter between error terms $\rho$. In addition, we included a random effect on the exclusion restriction variable given by $b_{i3}^S\sim N(0,0.2)$ assuming that the intracluster variation in the exclusion restriction effect is lower than the variation on other coefficient parameters effects. The $\rho$ parameter was given different values depending on the simulated missing mechanism (Table \ref{tab:scenarios}) .

As regards the error terms, they were bivariate normal distributed as:
\[\begin{pmatrix}
\epsilon^O_{i} \\
\epsilon^S_{i}
\end{pmatrix}\sim N_{n_i}\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\begin{pmatrix}
\sigma^2_i & \rho\sigma_i \\
\rho\sigma_i & 1
\end{pmatrix}\right)
\]
whose $\sigma_i^2$ is a random parameter across clusters distributed as $log(\sigma_i) \sim N(0,0.05)$.

@@

## Estimands
The estimands were the parameter coefficients of the outcome equation $\boldsymbol{\beta^O}=(\beta_{0}^O,\beta_{1}^O,\beta_{2}^O)^T$, with special emphasis on the treatment effect parameter $\beta^0_1$. We also report the estimated standard deviation from the covariance matrix of the random effects, i.e.,  $\sqrt{\psi^O_{00}}$,$\sqrt{\psi^O_{11}}$,$\sqrt{\psi^O_{22}}$.

@@

After the imputation procedure, for the scenarios with incomple continuos outcome, we estimated the following mixed linear effect model using the *lmer()* function from the *lme4* R package.[@bates2022]
$$\boldsymbol{y_{i}}=\boldsymbol{X_i^O}\boldsymbol{\beta_i^O}+\boldsymbol{\epsilon^O_{i}}$$
with $\beta_i^O=\beta^O+b_i^O$, $b_i^O\sim N(0,\Psi^O)$ and $\epsilon^O_{i} \sim N(0,\Sigma_i)$. In case of an incomplete binary outcome, we used the same matrix of predictors but on a generalized linear mixed model estimated with the *glmer()* function from the *lme4* R package. Then, we pooled the estimates of the $\beta_i^O$ and the variance of the random effect and residual errors of the multiple imputed datasets according to Rubin's rule[@rubin1987], over which we calculated the performance measures on the estimands.

@@

The 2l.Heckman imputation method is available in the micemd R package [@audigier2021] (as \textbf{mice.2l.2stage.heckman()}) and also on the github repository \url{https://github.com/johamunoz/Statsmed_Heckman} where you can also find all the codes accompanying this paper and a toy example that explains how to implement the method in mice.

@@

 ```{r rhocont,fig.align='center',fig.pos="!htb",fig.cap="Continuous incomplete variable by varing $\\rho$, where dashed lines depict the target performance criteria value",echo=FALSE,fig.keep="last",fig.height = 3.9, fig.width = 6}
plot_rho
 ```

Figure \ref{fig:rhocont} shows the results of simulations where the incomplete outcome was continuous. In the MAR scenario, i.e. when $\rho=0$, all imputation methods provided similar unbiased estimates of the coefficient parameters $\beta^O$ , but as $\rho$ increased, i.e. the mechanism became MNAR, the estimates for the complete-case analysis and the 2l.MAR imputation method became further away from the true value.

As expected, both Heckman-based imputation methods (1l.Heckman and 2l.Heckman) gave less biased estimates of the coefficient parameters $\beta^O$ in MNAR scenarios, but the estimates of the random effects parameters $\sqrt{\psi^O_{kk}}$ with $k =\{0,1,2\}$ in the 1l.Heckman method were the most unbiased. This could be explained by the fact that 1l.Heckman does not take into account any grouping information.

In terms of coverage, the 2l.Heckman imputation method gave the best coverage values in all $\rho$ scenarios with a coverage level close to the nominal 95\% value. Although 2l.Heckman was not properly a randomization-valid method [@buuren18c], as it was not unbiased and had a coverage above 95\% across all $\rho$ values, this method gave better results in terms of bias and coverage when compared to the other methods evaluated.  The 2l.Heckman method, also resulted in better $\beta^O$ estimates in terms of RMSE than the other methods evaluated, through a better compromise between bias and variance.

In particular, the 2l.Heckman method provides an advantage over the 1l.Heckman method on data sets with systematic omission, by including cluster information in the imputation model. This can be seen in the random effects parameter estimates $\sqrt{\psi^O_{kk}}$ of the 1l.Heckman method, which were more biased than those of the other methods where cluster information was included, i.e. 2l.MAR and 2l.Heckman.
